Spending too much time at the Galactic bar: Chaotic stream-fanning and the Ophiuchus stream

### 2016-01-04 ms.tex
ahhhh

The orbit fit and $N$-body simulations in S15a used a static, axisymmetric potential to represent the Milky Way potential, but it is well-known that the Galactic bulge contains a triaxial, bar-like structure several kpc in size \citep[e.g.,][]{blitz91, weinberg92, dwek95, wegg13}. Given the proximity of the stream to the center of the Galaxy, the time-dependent, triaxial potential of the Galactic bar must be taken into account when modeling the orbit of the Ophiuchus stream. The presence of a bar-like perturbation to the potential will change the orbit of the stream progenitor and the orbit structure in the inner galaxy \citep{portail15b, gajda15, zotos12}. Bar-like features can also introduce a significant number of chaotic orbits in their vicinity\citep{weinberg15}.

@@

To integrate orbits and to compute chaos indicators we must choose a gravitational potential model to represent the potential of the Milky Way. The key feature of the potential that we would like to capture is the time-dependence and triaxiality of the Galactic bar. Recent work has used stellar number counts of Red Clump giant stars in the Galactic bulge to constrain dynamical models of the bar \citep{portail15}. Measurements of the total mass of the bar feature from this study are largely consistent with past work \citep[e.g.,][]{wang12, XX}, however the measured pattern speed and present bar angle are significantly discrepant and this difference is not fully understood. We construct a parametrized potential model consisting of a triaxial, time-dependent (rotating) bulge component added to simple models for the disk and halo of the Milky Way. We describe below how we fix the parameters of the disk, halo, and bar or bulge component, but explore different choices for the time-dependence and orientation of the bar. We also define a static potential with a spherical bulge for comparison.

These potential models are meant to be representative rather than definitive. The uncertainty in the Milky Way potential within Galactocentric radii of $r\lesssim 4~{\rm kpc}$ and outside of $r\gtrsim 15~{\rm kpc}$ are large enough that trying to match the exact density distribution of the Ophiuchus stream is not a useful exercise. Instead, we consider qualitatively different potentials that allow us to isolate and study the affect of chaotic stream-fanning of tidal debris in the vicinity of the stream. \todo{KVJ -- is this better?}

@@

\begin{table}[ht]
\begin{center}
	\begin{tabular}{ c | c | c }
	         \toprule
	         Name & $\alpha$ [deg] & $\Omega_p$ [${\rm km}~{\rm s}^{-1}~{\rm kpc}^{-1}$] \\\toprule
		bar1 & 20 & 40\\
		bar2 & 20 & 50\\
		bar3 & 20 & 60\\
		bar4 & 25 & 40\\
		bar5 & 25 & 50\\
		bar6 & 25 & 60\\
		bar7 & 30 & 40\\
		bar8 & 30 & 50\\
		bar9 & 30 & 60\\
		\bottomrule
		\end{tabular}
	\caption{Present-day bar angle ($\alpha$) and pattern speed ($\Omega_p$) for the nine parameter pairs considered in this work. These values span the range of recent measurements from a variety of techniques \citep{dwek95,wang12,wang13, MOREMOREMORE}. \label{tbl:bar-specific}}
\end{center}
\end{table}

@@

We make one modification to this method based on the idea that the Ophiuchus stream progenitor has been fully-disrupted. We add an additional parameter to the stream generation routine to specify the time of disruption, $\tau_d$. At this time, we set the offset of the Lagrange point to 0: In terms of the parametrization in \citep{fardal14}, we set $k_r = 0$ and $k_{vt}=0$ but preserve the dispersion in the release radius and velocity. Any star released after $\tau_d$ is released with a small dispersion around the progenitor orbit but no offset. This mass-loss history is intended to mimic the expected gradual evaporation of a globular cluster over a tidally-limited boundary (i.e. driven by two-body relaxation and gravitational shocks over many Gigayears) with final disruption likely to occur once the tidal boundary is less than the core radius of the cluster. The physics of this disruption is not followed exactly but rather the disruption rate and final disruption time are set by matching the observed density along the stream. \todo{APW: check that last statement -- can we be more precise?}

@@

For the posterior samples in each potential, we also compute the maximum Lyapunov exponent (MLE, $\lyapexp$) and corresponding Lyapunov \emph{time} ($\lyapt = 1/\lyapexp$) to assess whether each orbit is chaotic. For strongly chaotic orbits, the Lyapunov time is still an appropriate indicator of chaos and of the timescale over which chaos is important for tidal debris \citep{apw15-chaos}. Figure~\ref{fig:lyapunov-hist} shows distributions of Lyapunov times for orbits drawn from the posterior distributions from orbit fitting in each potential model. All orbits in the static potential have Lyapunov times $\lyapt > 20~{\rm Gyr}$ and we consider them to be regular (no panel is shown for these orbits). All orbits sampled from each barred potential are strongly chaotic with Lyapunov times that range from $\lyapt \approx 400--1100~{\rm Myr}$. It is clear from these panels that the orbits around Ophiuchus are generally more strongly chaotic (have lower Lyapunov times) for larger pattern speeds.

@@

Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} show the final particle positions and line-of-sight velocities in heliocentric coordinates for the maximum likelihood mock streams (grey points) in each potential. Vertical, dashed lines show the approximate extent of the part of the stream visible in main-sequence stars (excluding the BHB stars from S15b). There are a few interesting features to note from these panels: 
\begin{enumerate}
	\item Even in the static potential (Figure~\ref{fig:mockstream0}, leftmost column), there is a slight decrease in the density for the model stream towards higher Galactic longitudes, $l$. This is a projection effect: The portion of the stream at larger $l$ is closer points almost directly towards the Sun so that the debris covers a larger area. This is more clearly seen in Figure~\ref{fig:mockstreamxyz}, which shows $x$-$z$ projections of the star particles in Galactocentric Cartesian coordinates (grey) along with the position of the Sun (symbol at $(x,z)=(-8.3,0)~{\rm kpc}$) and the ``window'' of the Heliocentric, sky-position plots of Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} (shown as blue lines).
	\item The density of the mock stream in the static potential decreases slowly rather than abruptly as is observed. This is more clearly seen in terms of the \todo{linear density along the stream, shown in Figure~\ref{fig:lineardensity}}. .... \todo{Waiting on Branimir!}
	\item \todo{4 BHBs beyond end of stream never fit in pos/vel for static case, but do in some bar potentials?}
	\item For the barred potentials, the stream morphology is very sensitive to the properties of the bar (especially the pattern speed) and to the initial conditions of the orbit. In addition to generating mock streams on the mean orbits, we have run similar simulations for random draws from the posterior distributions over orbit fits. We have found that the morphology can vary significantly between different samples (because these are strongly chaotic orbits), but the overall characteristics remain similar: along more strongly chaotic orbits, the debris ``fans'' more and the apparent dense part of the model streams is shorter (see, for example, Figure~\ref{fig:mockstreamxyz}, panels bar2 vs. bar3).
\end{enumerate}

@@

The current kinematic data for the Ophiuchus BHB stars suggests that the stream is sensitive to the potential of the bar, but with better measurements of the velocities and a larger sample of member stars we would be able to constrain the bar model parameters. Current measurements of the pattern speed, angle, and structure of the Milky Way's bulge and bar are largely discrepant \citep[e.g.,][]{wang12, wang13, antoja14, wegg?? is wegg appropriate here}. Most of the methods that infer these quantities rely on modeling the density or kinematics of stars at low Galactic latitudes and must therefore handle challenges with completeness and dust extinction. The Ophiuchus stream offers a unique opportunity to independently measure these quantities by modeling the density and kinematics of stars associated with the stream.

@@

\todo{Still working on what to say here - suggestions welcome!}
If the Ophiuchus stream formed from a globular cluster on a chaotic orbit and we happen to be witnessing its final demise, what does this imply about the population of clusters that have already been fully disrupted? For a given model for the Galactic bar, we can study the orbit structure in the inner Milky Way and, in particular, study the fraction of strongly chaotic orbits that would lead to chaotic fanning or fast dispersal of tidal debris \citep{recent work by zotos or those greek peeps}. The existence of strongly chaotic orbits in this region would limit the expected number of cold stellar streams in the inner Galaxy. Comparing the observed numbers with expectations ...

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mockstream0}
\caption{ Sky position, distance, and line-of-sight velocity in Heliocentric, Galactic coordinates for star particles (grey points) from mock streams generated on the mean orbits in each potential. Vertical, dashed lines show the approximate extent of the densest part of the stream visible in main-sequence stars \citep[the segment originally detected in ][]{bernard14}.\todo{make top panels density maps instead}}
\label{fig:mockstream0}
\end{center}
\end{figure*}

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/surface-density}
\caption{ Surface density of mock stream star particles in each potential. Contours are spaced logarithmically from $10^{-2}$ to $10$ particles per sq. deg---that is, each color represents a factor of 10 difference in surface density. In the static potential (top left), the density remains high along the center of the stream, but for some of the barred potentials the density drops sharply because of chaotic stream-fanning.}
\label{fig:surfacedensity}
\end{center}
\end{figure*}

@@

For a given set of initial conditions ($\bs{w}_0$), we compute a model orbit as follows: (1) transform the initial conditions to Galactocentric coordinates, where we take the position and velocity of the Sun to be 
\begin{align}
	(x,y,z)_\odot &= (-8,0,0)~{\rm kpc}\\
	(v_x,v_y,v_z)_\odot &= (-11.1, 220+24, 7.25)~{\rm km}~{\rm s}^{-1}
\end{align}
\citep{8kpc 	ref??, schonrich10, bovy12}, (2) integrate the orbit forward and backward by $t_f$ and $t_b$, respectively, in the potential $\Phi$, (3) transform all orbit points (time-steps) back to observed coordinates, and (4) define interpolating functions for each coordinate as a function of stream longitude, $\phi_1$, using cubic splines---e.g., functions $\widetilde{\phi}_{2}(\phi_1)$, $\widetilde{\DM}(\phi_1)$, $\widetilde{\mu_l}(\phi_1)$, $\widetilde{\mu_b}(\phi_1)$, $\widetilde{v_r}(\phi_1)$. These functions let us compute the predicted values of each of these coordinates at the longitudes of each observed star, $\phi_{1,i}$.

>>>>

The orbit fit and $N$-body simulations in S15a used a static, axisymmetric potential to represent the Milky Way potential, but it is well-known that the Galactic bulge contains a triaxial, bar-like structure several kpc in size \citep[e.g.,][]{blitz91, weinberg92, dwek95, wegg13}. Given the proximity of the stream to the center of the Galaxy, the time-dependent, triaxial potential of the Galactic bar must be taken into account when modeling the orbit of the Ophiuchus stream. The presence of a bar-like perturbation to the potential will change the orbit of the stream progenitor and the orbit structure in the inner galaxy \citep{zotos12, portail15b, gajda15}. Bar-like features can also introduce a significant number of chaotic orbits in their vicinity \citep{weinberg15}.

@@

To integrate orbits and to compute chaos indicators we must choose a gravitational potential model to represent the potential of the Milky Way. The key feature of the potential that we would like to capture is the time-dependence and triaxiality of the Galactic bar. Recent work has used stellar number counts of Red Clump giant stars in the Galactic bulge to constrain dynamical models of the bar \citep{portail15}. Measurements of the total mass of the bar feature from this study are largely consistent with past work \citep[e.g.,][]{wang12}, however the measured pattern speed and present bar angle are significantly discrepant and this difference is not fully understood. We construct a parametrized potential model consisting of a triaxial, time-dependent (rotating) bulge component added to simple models for the disk and halo of the Milky Way. We describe below how we fix the parameters of the disk, halo, and bar or bulge component, but explore different choices for the time-dependence and orientation of the bar. We also define a static potential with a spherical bulge for comparison.

These potential models are meant to be representative rather than definitive. The uncertainty in the Milky Way potential within Galactocentric radii of $r\lesssim 4~{\rm kpc}$ and outside of $r\gtrsim 15~{\rm kpc}$ are large enough that trying to match the exact density distribution of the Ophiuchus stream is not a useful exercise. Instead, we consider qualitatively different potentials that allow us to isolate and study the affect of chaotic stream-fanning of tidal debris in the vicinity of the stream.

@@

\begin{table}[ht]
\begin{center}
	\begin{tabular}{ c | c | c }
	         \toprule
	         Name & $\alpha$ [deg] & $\Omega_p$ [${\rm km}~{\rm s}^{-1}~{\rm kpc}^{-1}$] \\\toprule
		bar1 & 20 & 40\\
		bar2 & 20 & 50\\
		bar3 & 20 & 60\\
		bar4 & 25 & 40\\
		bar5 & 25 & 50\\
		bar6 & 25 & 60\\
		bar7 & 30 & 40\\
		bar8 & 30 & 50\\
		bar9 & 30 & 60\\
		\bottomrule
		\end{tabular}
	\caption{Present-day bar angle ($\alpha$) and pattern speed ($\Omega_p$) for the nine parameter pairs considered in this work. These values span the range of recent measurements from a variety of techniques \citep{dwek95, wang12,wang13,wegg13}. \label{tbl:bar-specific}}
\end{center}
\end{table}

@@

We make one modification to this method based on the idea that the Ophiuchus stream progenitor has been fully-disrupted. We add an additional parameter to the stream generation routine to specify the time of disruption, $\tau_d$. At this time, we set the offset of the Lagrange point to 0: In terms of the parametrization in \citep{fardal14}, we set $k_r = 0$ and $k_{vt}=0$ but preserve the dispersion in the release radius and velocity. Any star released after $\tau_d$ is released with a small dispersion around the progenitor orbit but no offset. This mass-loss history is intended to mimic the expected gradual evaporation of a globular cluster over a tidally-limited boundary (i.e. driven by two-body relaxation and gravitational shocks over many Gigayears) with final disruption likely to occur once the tidal boundary is less than the core radius of the cluster. The physics of this disruption is not followed exactly but rather the disruption rate and final disruption time are set by \todo{the hypothesis that the most recent shock ...}

@@

For the posterior samples in each potential, we also compute the maximum Lyapunov exponent (MLE, $\lyapexp$) and corresponding Lyapunov \emph{time} ($\lyapt = 1/\lyapexp$) to assess whether each orbit is chaotic. For strongly chaotic orbits, the Lyapunov time is still an appropriate indicator of chaos and of the timescale over which chaos is important for tidal debris \citep{apw15-chaos}. Figure~\ref{fig:lyapunov-hist} shows distributions of Lyapunov times for orbits drawn from the posterior distributions from orbit fitting in each potential model. All orbits in the static potential have Lyapunov times $\lyapt > 20~{\rm Gyr}$ and we consider them to be regular (no panel is shown for these orbits). All orbits sampled from each barred potential are strongly chaotic with Lyapunov times that range from $\lyapt \approx 400$--$1100~{\rm Myr}$. It is clear from these panels that the orbits around Ophiuchus are generally more strongly chaotic (have lower Lyapunov times) for larger pattern speeds.

@@

Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} show the final particle positions and line-of-sight velocities in heliocentric coordinates for the maximum likelihood mock streams (grey points) in each potential. Vertical, dashed lines show the approximate extent of the part of the stream visible in main-sequence stars (excluding the BHB stars from S15b). There are a few interesting features to note from these panels: 
\begin{enumerate}
	\item Even in the static potential (Figure~\ref{fig:mockstream0}, leftmost column), there is a slight decrease in the density for the model stream towards higher Galactic longitudes, $l$. This is a projection effect: The portion of the stream at larger $l$ is closer points almost directly towards the Sun so that the debris covers a larger area. This is more clearly seen in Figure~\ref{fig:mockstreamxyz}, which shows $x$-$z$ projections of the star particles in Galactocentric Cartesian coordinates (grey) along with the position of the Sun (symbol at $(x,z)=(-8.3,0)~{\rm kpc}$) and the ``window'' of the Heliocentric, sky-position plots of Figures~\ref{fig:mockstream0}--\ref{fig:mockstream1} (shown as blue lines).
	\item The density of the mock stream in the static potential decreases slowly rather than abruptly as is observed. This is more clearly seen in terms of the \todo{linear density along the stream, shown in Figure~\ref{fig:lineardensity}}. .... \todo{Waiting on Branimir!}
	\item The four high-dispersion BHB stars beyond the end of the stream (from S15b) don't match in position and velocity with the particle distribution from even the maximum likelihood stream model in the static potential. However, in some barred potentials the chaotic evolution of the stream stars can lead to over-densities of stars with an increased positional dispersion and significantly discrepant velocities (e.g., bar8).
	\item None of the stream models produce an appreciable density of stars with line-of-sight velocities near the S15b BHB star with the largest velocity ($\approx 320~\kms$). This star is either an interesting Ophiuchus member star or is associated with some other kinematic substructure.
	\item For the barred potentials, the stream morphology is very sensitive to the properties of the bar (especially the pattern speed) and to the initial conditions of the orbit. We have found that the morphology can vary significantly between nearby orbits in the same potential model (because these are strongly chaotic orbits), but the overall characteristics remain similar: along more strongly chaotic orbits, the debris ``fans'' more and the apparent dense part of the model streams is shorter (see, for example, Figure~\ref{fig:mockstreamxyz}, panels bar3 vs. bar8).
\end{enumerate}

@@

The current kinematic data for the Ophiuchus BHB stars suggests that the stream is sensitive to the gravitational potential of the Milky Way bar---with better measurements of the velocities and a larger sample of member stars we will constrain parameters for the bar model. Current measurements of the pattern speed, angle, and structure of the Milky Way's bulge and bar are largely discrepant \citep[e.g.,][]{wang12, wang13, wegg13, antoja14}. Most of the methods that infer these quantities rely on modeling the density or kinematics of stars at low Galactic latitudes and must therefore handle challenges with completeness and dust extinction. The Ophiuchus stream offers a unique opportunity to independently measure these quantities by modeling the density and kinematics of stars associated with the stream.

@@

If the Ophiuchus stream formed from a globular cluster on a chaotic orbit and we happen to be witnessing its final demise, what does this imply about the population of clusters that have already been fully disrupted? For a given model for the Galactic bar, we will study the orbit structure in the inner Milky Way and, in particular, study the fraction of strongly chaotic orbits that would lead to chaotic fanning or fast dispersal of tidal debris \citep{recent work by zotos or those greek peeps}. The existence of strongly chaotic orbits in this region would limit the expected number of cold stellar streams in the inner Galaxy and enhance the rate of mixing of the debris. Gaia will .... compare amount of substructure with expectations ..... \todo{Quote Gnedin and Ostriker, Murali and Weinberg}

@@

\begin{figure*}[p]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mockstream0}
\caption{ Sky position, distance, and line-of-sight velocity in Heliocentric, Galactic coordinates for star particles (contours or grey points) from the maximum-likelihood mock streams in each potential. Top panels show surface density of mock stream star particles in each potential. Contours are spaced logarithmically from $10^{-2}$ to $10$ particles per sq. deg---that is, each color represents a factor of 10 difference in surface density. In the static potential (top left), the density remains high along the center of the stream, but for some of the barred potentials the density drops sharply because of chaotic stream-fanning. Vertical, dashed lines show the approximate extent of the densest part of the stream visible in main-sequence stars \citep[the segment originally detected in ][]{bernard14}.}
\label{fig:mockstream0}
\end{center}
\end{figure*}

@@

%\begin{figure*}[p]
%\begin{center}
%\includegraphics[width=\textwidth]{figures/surface-density}
%\caption{ Surface density of mock stream star particles in each potential. Contours are spaced logarithmically from $10^{-2}$ to $10$ particles per sq. deg---that is, each color represents a factor of 10 difference in surface density. In the static potential (top left), the density remains high along the center of the stream, but for some of the barred potentials the density drops sharply because of chaotic stream-fanning.}
%\label{fig:surfacedensity}
%\end{center}
%\end{figure*}

@@

For a given set of initial conditions ($\bs{w}_0$), we compute a model orbit as follows: (1) transform the initial conditions to Galactocentric coordinates, (2) integrate the orbit forward and backward by $t_f$ and $t_b$, respectively, in the potential $\Phi$, (3) transform all orbit points (time-steps) back to observed coordinates, and (4) define interpolating functions for each coordinate as a function of stream longitude, $\phi_1$, using cubic splines---e.g., functions $\widetilde{\phi}_{2}(\phi_1)$, $\widetilde{\DM}(\phi_1)$, $\widetilde{\mu_l}(\phi_1)$, $\widetilde{\mu_b}(\phi_1)$, $\widetilde{v_r}(\phi_1)$. These functions let us compute the predicted values of each of these coordinates at the longitudes of each observed star, $\phi_{1,i}$.
